Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-08-12 supplement.tex
added figures

@@

In our primary model fits, we used the Mende in Sierra Leone (MSL) and the
British from England and Scotland (GBR). We tested the same models against a
different West African population (the Yoruba from Ibidan, Nigeria, YRI), and
two different European populations (Western Europeans from Utah, CEU, and
Toscani in Italy, TSI). The CEU are closely related to GBR and are thus likely
to be an equivalently close proxy for European ancestry in African individuals
with admixed European ancestry. The TSI are likely more distantly related to
the source of this European ancestry. We chose not to replace the GBR
non-European Eurasian populations, as these would be poorer proxies for early
Holocene and recent colonial migration from the Near East and Europe back to
Africa, and it is unclear how such a model should be interpreted. We therefore
refit each of our best-fit demographic models using the following alternate
population sets: (Nama, YRI, Gumuz, Oromo/Amhara, and CEU), and (Nama, MSL,
Gumuz, Oromo/Amhara, and TSI).

@@

Because CEU and GBR are genetically closely related, we also tested the
consistency of inferred parameters when replacing GBR with Toscani from Italy,
a more distantly related European population set than GBR. Most parameters in
each of the tested models were consistent with parameters inferred using GBR.
The only large difference was found in fitting the continuous migration model,
Nama divergence was $\approx200$ka, rather than $135$ka. Migration rates during
this early period of the model also shifted to larger values between the stem
populations, demonstrating that divergence times and migration rates are
confounded parameters in this model. Notably, the multiple merger model (which
provides a better fit to the data) was inferred to have consistent divergence
and merger times between the models fit using GBR and TSI, with ancestral
populations ``reshuffling'' between 100 and 120ka.

\aprcomment{Short paragraph here explaining the discrepancy in the TSI fit
continuous migration model}

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[width=5in]{figures/supp-toy-models}
    \caption{The impact of model specification on coalescence rate and inference.
    We show five simplified demographic models that differ in demographic events, 
    occurring between 200ka and 100ka, that mark the transition between a single 
    ancestral population and two daughter populations. A) A clean split 100ka, 
B) A clean split 200ka, C) population growth at 200ka followed by split 100ka, 
D) split 200ka followed by migration until 100ka, and E) split 200ka followed by 
merger 105ka and a new split 100ka. The left column shows model representation, 
the middle column shows instantaneous coalescence rates for pairs of lineages from the
same daughter population, the right column shows instantaneous coalescence rates for
pairs of lineages coming from distinct populations.  
        \label{fig:toymodels}
    }
    \end{figure}

>>>>

\usepackage{caption}
\usepackage{subcaption}
\usepackage{graphicx}

@@

In our primary model fits, we used the Mende in Sierra Leone (MSL) and the
British from England and Scotland (GBR). We tested the same models against a
different West African population (the Yoruba from Ibidan, Nigeria, YRI), and
two different European populations (Western Europeans from Utah, CEU, and
Toscani in Italy, TSI). The CEU are closely related to GBR and are thus likely
to be an equivalently close proxy for European ancestry in African individuals
with admixed European ancestry. The TSI are likely more distantly related to
the source of this European ancestry. We chose not to replace the GBR
by non-European Eurasian populations, as these would be poorer proxies for early
Holocene and recent colonial migration from the Near East and Europe back to
Africa, and it is unclear how such a model should be interpreted. We therefore
refit each of our best-fit demographic models using the following alternate
population sets: (Nama, YRI, Gumuz, Oromo/Amhara, and CEU), and (Nama, MSL,
Gumuz, Oromo/Amhara, and TSI).

@@

Because CEU and GBR are genetically closely related, we also tested the
consistency of inferred parameters when replacing GBR with Toscani from Italy,
a more distantly related European population set than GBR. Most parameters in
each of the tested models were consistent with parameters inferred using GBR.
The only large difference was found in fitting the continuous migration model,
Nama divergence was $\approx200$ka, rather than $135$ka. 
Our original model using the dataset with GBR
underestimated both $\sigma_{Dz}$ and $\sigma_{\pi_2}$ compared to data
(Figure~\ref{fig:supp-continuous-migration-fits}), and as a result provided a poorer 
fit to the data than the multiple merger model. 
When fitting this same model to the TSI, the model provided a better fit to the $\sigma_{Dz}$ statistics but
a worse fit to the $\sigma_{\pi_2}$ statistics (and still provided a
significantly worse fit to the data compared to the multiple merger model).
Migration rates during
this early period of the model also shifted to larger values between the stem
populations, demonstrating that divergence times and migration rates are
confounded parameters in this model. 
These differences in inferred parameters are likely driven by a combination of factors: the continuous
migration model mis-specifies historical dynamics in the range of 100-200ka;
the TSI are an inappropriate proxy for recent European admixture in East and
South Africa; the TSI may also carry recent North African ancestry, which
is not captured in our model. Finally, there is an amount of confounding between divergence time
and migration rates.

By contrast, the multiple merger model (which
provides a better fit to the data) was inferred to have consistent divergence
and merger times between the models fit using GBR and TSI, with ancestral
populations ``reshuffling'' between 100 and 120ka.

Across all validations to alternative datasets described in this section, model
fits were largely robust and matched those shown in the main text
(Figure~\ref{fig:best-fit-models})---the single outlier in these validations
was the continuous migration model fit to the dataset that included TSI instead
of GBR. We conclude that while there remain important issues

Together, this highlights the sensitivity of
complex demographic inference to the choices of populations to include, as well
as to choices of model specification.
\aprcomment{not sure about these last two sentences}

@@

\begin{figure}

    \includegraphics{figures/toy_models/toy_model_combo}

    \caption{The impact of model specification on coalescence rate and inference.
    We show five simplified demographic models that differ in demographic events, 
    occurring between 200ka and 100ka, that mark the transition between a single 
    ancestral population and two daughter populations. A) A clean split 100ka, 
B) A clean split 200ka, C) population growth at 200ka followed by split 100ka, 
D) split 200ka followed by migration until 100ka, and E) split 200ka followed by 
merger 105ka and a new split 100ka. The left column shows model representation, 
the middle column shows instantaneous coalescence rates for pairs of lineages from the
same daughter population, the right column shows instantaneous coalescence rates for
pairs of lineages coming from distinct populations.  
        \label{fig:toymodels}
    }

    \end{figure}
